The figure below shows the raw choice data for the participants that have been excluded.
An online app where the full dataset can be interactively explored is hosted at \mbox{\href{https://fdabl.shinyapps.io/MCMCP/}{https://fdabl.shinyapps.io/MCMCP/}}. All materials are available at \mbox{\href{https://github.com/fdabl/MCMCP}{https://github.com/fdabl/MCMCP}}.

The participant on the top left was excluded due to a bias towards small values, while three other participants were excluded due to a bias towards large values. The fifth participant who was excluded did not seem to understand the experiment, unsystematically alternating between images with a higher and lower number of red dots.

```{r, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, echo = FALSE, warning = FALSE)
```

```{r}
library('dplyr')
library('ggmcmc')
library('stringr')
library('ggplot2')
library('gridExtra')
source('../exploring/prepare_exploring.R') # provides the data frame d

# keep the participants who have been removed
to_remove = read.csv('../exploring/ids.csv') %>% filter(!(number %in% c(10, 11, 15, 22, 34)))
d = filter(d, !(id %in% to_remove$id)) %>% 
  mutate(language = as.factor(str_to_lower(language))) %>% tbl_df

# TODO: write a function that quantifiers the distance to the mean
```

```{r, fig.height=18, fig.width = 16, fig.cap="Raw data for the excluded participants. Blue line indicates the mode of the subjective probability distribution for that specific quantifier (estimated without excluded participants)."}
ids = unique(d$id)

plot_data = function(id) {
  test = melt_data(d, id) %>%
    mutate(Predictions = ifelse(chose_higher, pred_distance, 1 - pred_distance),
           mode = mode_distance)
  
  p = ggplot(test, aes(x = trial, y = value)) +
        geom_point(size = 3, color = 'grey30') +
      
        geom_path(data = filter(test, choice == 'number_chosen'), size = 1.2) +
        xlab('Trial') + ylab('Number') +
        geom_hline(yintercept = c(108, 216, 324, 432), linetype = 'dotted') +
        geom_hline(aes(yintercept = mode), color = 'blue') +
        scale_y_continuous(breaks = scales::pretty_breaks(n = 15),
                           limits = c(0, 432)) + theme_bw() +
        theme(plot.title = element_text(size = 24),
              axis.title = element_text(size = 22))
  
  p + facet_wrap(~ quantifier)
}

grid.arrange(plot_data(ids[1]), plot_data(ids[2]),
             plot_data(ids[3]), plot_data(ids[4]), plot_data(ids[5]))
```